---
title: "Covariate Selection and What Actually Ran"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariate Selection and What Actually Ran}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, message = FALSE}
library(estimatrTools)
```

Covariate adjustment in a randomized experiment is supposed to buy precision
without costing unbiasedness. Choosing the covariates by hand invites
specification search; choosing all of them costs degrees of freedom, and under
the Lin estimator costs them once per arm. This package selects them by LASSO
and then fits the selected set, and it keeps a record of what it actually did,
which turns out to matter more than it sounds.

## Selection

```{r data}
set.seed(20260725)
n <- 800

dat <- data.frame(
  Z          = rbinom(n, 1, 0.5),
  X_prog     = rnorm(n),   # predicts the outcome
  X_het      = rnorm(n),   # predicts the outcome only among the treated
  X_noise_1  = rnorm(n),
  X_noise_2  = rnorm(n)
)
dat$Y <- 0.3 * dat$Z + 1.5 * dat$X_prog + 2 * dat$Z * dat$X_het + rnorm(n)

candidates <- ~ X_prog + X_het + X_noise_1 + X_noise_2
lasso_select_covariates(Y ~ Z, candidates, data = dat)
```

Both real covariates are found and the noise is not. `X_het` is the interesting
one: it predicts the outcome only in the treated arm, so a pooled regression of
`Y` on the candidates would barely notice it. Selection runs the outcome
equation *within each arm*, which is what makes it appropriate for an estimator
that fits a separate slope per arm.

The second family of regressions is on the arm indicators, the double-selection
step of Belloni, Chernozhukov and Hansen (2014). It retains covariates that
predict assignment even when they barely predict the outcome, which is the case
where omitting one does the most damage.

## Three estimators

```{r fits}
lin <- lm_lin_lasso(Y ~ Z, candidates, data = dat)
add <- lm_robust_lasso(Y ~ Z, candidates, data = dat)

broom::tidy(lin)[1:2, c("term", "estimate", "std.error")]
broom::tidy(add)[1:2, c("term", "estimate", "std.error")]
```

`lm_lin_lasso()` fits a separate covariate slope in each arm, with covariates
centered so the treatment coefficient still estimates the average treatment
effect. `lm_robust_lasso()` enters them additively, assuming a common slope.
That assumption buys degrees of freedom, which is what makes it the better
choice when arms are small, but when it is wrong the treatment coefficient is no
longer guaranteed to be consistent for the ATE. Lin (2013) recommends the
interacted form as the default for that reason. The choice is yours to make, not
the package's, which is why they are separate functions rather than an argument.

`lm_moderator_lasso()` fits `Y ~ Z * moderator` with the selected covariates entered
additively alongside, deliberately not interacted with treatment or moderator.

## What actually ran

Adjustment is not always possible. `lm_lin` cannot be fitted with no covariates,
and an adjusted fit can come back degenerate when the selected set is nearly
collinear within an arm. In either case these functions return an unadjusted
fit, which for a binary treatment is the difference in means.

That is usually the right response. It also changes what the estimate is, so it
is recorded.

```{r fallback}
noise_only <- data.frame(Z = rbinom(n, 1, 0.5), X_noise = rnorm(n))
noise_only$Y <- 0.3 * noise_only$Z + rnorm(n)

fit <- lm_lin_lasso(Y ~ Z, ~ X_noise, data = noise_only)

adjustment(fit)
fallback_reason(fit)
```

**This is common, not exceptional.** In one meta-analysis pipeline, 38% of
roughly a thousand `lm_lin_lasso()` calls fell back, nearly all because LASSO
selected nothing. In those cases an estimate described in the write-up as
"Lin-adjusted" is an unadjusted difference in means. That is not wrong, and it
is what the estimator is defined to do, but a reader comparing adjusted and
unadjusted estimates should know that a large share of the pairs are identical
by construction rather than by agreement.

Three accessors, depending on what you still hold. For one fit:

```{r one-fit}
c(adjustment = adjustment(fit), reason = fallback_reason(fit))
```

For a list of fits:

```{r fit-list}
fits <- list(
  informative = lm_lin_lasso(Y ~ Z, candidates, data = dat),
  noise_only  = lm_lin_lasso(Y ~ Z, ~ X_noise, data = noise_only)
)

fallback_summary(fits)
```

And for fits that no longer exist. This is the case that matters in practice,
because the common shape is to build a fit inside `map()`, summarize it, and
drop the model column:

```{r log}
reset_fallback_log()

for (i in 1:5) {
  d <- if (i %% 2 == 0) noise_only else dat
  cand <- if (i %% 2 == 0) ~ X_noise else candidates
  invisible(lm_lin_lasso(Y ~ Z, cand, data = d))   # fit is discarded
}

fallback_log()
```

The fits are gone; the record is not. The question a pipeline should be able to
answer is what share of its estimates were actually adjusted:

```{r rate}
table(fallback_log()$adjustment)
```

Call `reset_fallback_log()` at the top of a pipeline so the log describes that
run. Turn it off with `options(estimatrTools.log = FALSE)`.

## References

Belloni, A., Chernozhukov, V., and Hansen, C. (2014). Inference on treatment
effects after selection among high-dimensional controls. *Review of Economic
Studies*, 81(2), 608-650.

Lin, W. (2013). Agnostic notes on regression adjustments to experimental data:
reexamining Freedman's critique. *Annals of Applied Statistics*, 7(1), 295-318.
